Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage

### 2023-10-18 ms/ms.Rmd
added results and figure for fem model

---
# This template is largely functionally equivalent to that for elsevier
# The two fields that are not in the elsevier format are:
#      `corresponding_author` and `acknowledgements`
title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage
#date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: jwatts@colgate.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: Colgate University
  - name: Graham Dow
    email: graham.dow@usys.ethz.ch
    affiliation: ETH
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: University of California Davis
    footnote: 1 # Footnote using reference
  - name: Christopher D. Muir
    email: cdmuir@hawaii.edu
    affiliation: University of Hawaii at Manoa
    footnote: 1 # Footnote using reference
address:
  - code: Colgate University
    address: 13 Oak Drive, Hamilton, NY 13346
  - code: University of Hawaii at Manoa
    address: 2500 Campus Rd, Honolulu, HI 96822
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: \\href{mailto:cat@example.com}{cat@example.com}"
abstract: |
  This is the abstract.

  It consists of two paragraphs.
acknowledgements: |
  This is an acknowledgement.

  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage{booktabs} # for tables
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}

@@

# print(xtable::xtable(model_var, caption = paste0("The parameter range of model variables tested for their effect on coordination advantage (\\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation. We used regularly spaced values within each range and simulated across all combinations. Here we converted model units to more conventional units (e.g. m to $\\mu$m). ", symbol_descriptions), label = "tab:model_var"),
#       comment = FALSE, include.rownames = FALSE, type = "latex")

@@

@@

Stomata are expensive. A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata. Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum. This study provides evidence that stomata in *Arbidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}). However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}). Additionally, when grown in high light environments, *A. thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (\autoref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion. In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface. This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (\autoref{fig:length-area}). These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.

@@

>>>>

---
# This template is largely functionally equivalent to that for elsevier
# The two fields that are not in the elsevier format are:
#      `corresponding_author` and `acknowledgements`
title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage
#date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: jwatts@colgate.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: Colgate University
  - name: Graham Dow
    email: graham.dow@usys.ethz.ch
    affiliation: ETH
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: University of California Davis
    footnote: 1 # Footnote using reference
  - name: Christopher D. Muir
    email: cdmuir@hawaii.edu
    affiliation: University of Hawaii at Manoa
    footnote: 1 # Footnote using reference
address:
  - code: Colgate University
    address: 13 Oak Drive, Hamilton, NY 13346
  - code: University of Hawaii at Manoa
    address: 2500 Campus Rd, Honolulu, HI 96822
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: \\href{mailto:cat@example.com}{cat@example.com}"
abstract: |
  This is the abstract.
  It consists of two paragraphs.
acknowledgements: |
  This is an acknowledgement.
  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}

@@

@@

## Little benefit of coordinated stomatal arrangement

We used the finite element method (FEM) to model CO$_2$ diffusion within the leaf and photosynthesis as a 2-D porous medium. Across all realistic parts of parameter space, the coordination advantage is much less than 0.01 (\autoref{fig:model_summary}). For reference, a log-response of ratio is 0.01 is approximately 1%. The only exception was for thin leaves ($T_\text{leaf} = 100~\mu \text{m}$) with few stomata ($U = 338~\mu \text{m}$, which corresponds to a stomatal density of $\approx 10~\text{mm}^{-2}$), where lateral diffusion is major constraint on CO$_2$ supply. However, such thin leaves with so few stomata are uncommon among C$_3$ plants (some CAM plants have low stomatal density [@males_stomatal_2017]). In other areas of parameter space, lateral diffusion limitations were small relative to those along the ab-adaxial axis (see \autoref{fig:model_example} for a representative model solution).

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. `r symbol_descriptions`.}
\label{fig:model_summary}
\end{figure}

@@

Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017]. A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata. Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum. This study provides evidence that stomata in *Arabidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}). However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}). Additionally, when grown in high light environments, *A. thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (\autoref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion. In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface. This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (\autoref{fig:length-area}). These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.

@@

<!-- Placeholders for supplemental figures

\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\renewcommand\theequation{S\arabic{equation}}    
\setcounter{figure}{0}    
\setcounter{table}{0}    
\setcounter{equation}{0}

\begin{figure}
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:model_example}
\end{figure}
-->
